---
title: "HW 2: From correlation to linear mixed-effect models. Assignment sheet"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(vcd)
```

## 1. Vowel reduction in Russian
Pavel Duryagin ran an experiment on perception of vowel reduction in Russian language. The dataset `shva` includes the following variables:  
_time1_ - reaction time 1  
_duration_ - duration of the vowel in the stimuly (in milliseconds, ms)  
_time2_ - reaction time 2  
_f1_, _f2_, _f3_ - the 1st, 2nd and 3rd formant of the vowel measured in Hz (for a short introduction into formants, see [here](https://home.cc.umanitoba.ca/~krussll/phonetics/acoustic/formants.html))  
_vowel_ - vowel classified according the 3-fold classification (_A_ - _a_ under stress, _a_ - _a/o_ as in the first syllable before the stressed one, _y_ (stands for shva) - _a/o_ as in the second etc. syllable before the stressed one or after the stressed syllable, cf. _g_[_y_]_g_[_a_]_t_[_A_]_l_[_y_] _gogotala_ `guffawed').  
In this part, we will ask you to analyse correlation between f1, f2, and duration.
The dataset is available [https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/data/duryagin_ReductionRussian.txt](here).

### 1.0 Read the data from file to the variable `shva`.
```{r 1.0}
shva = read.delim("~/Yandex.Disk.localized/Учёба/ВШЭ/Лингвистическая теория и описание языка/Data Analysis/duryagin_ReductionRussian.txt")
shva
```

### 1.1 Scatterplot `f1` and `f2` using `ggplot()`. 
Design it to look like the [following](https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/img/duryagin1.png).
```{r 1.1}
ggplot(shva, aes(x=f2, y=f1, color=vowel)) + geom_point() + scale_y_continuous(trans = "reverse") + scale_x_continuous(trans = "reverse") + theme(legend.position="none") + ggtitle("f2 and f1 of the reduced and stressed vowels") + labs(caption="Data from Duryagin 2018")
```

### 1.2 Plot the boxplots of `f1` and `f2` for each vowel using `ggplot()`.
Design it to look like [this](https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/img/duryagin2.png) and [this](https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/img/duryagin3.png).
```{r 1.2}
# f1 boxplot
ggplot(shva, aes(x=vowel, y=f1, fill=vowel)) + geom_boxplot() + ggtitle("f1 distribution in each vowel") + labs(caption="Data from Duryagin 2018") + theme(legend.position="none", axis.title.y=element_blank()) + coord_flip()

# f2 boxplot
ggplot(shva, aes(x=vowel, y=f2, fill=vowel)) + geom_boxplot() + ggtitle("f2 distribution in each vowel") + labs(caption="Data from Duryagin 2018") + theme(legend.position="none", axis.title.y=element_blank()) +coord_flip()
```

### 1.3 Which `f1` can be considered outliers in _a_ vowel?
We assume outliers to be those observations that lie outside 1.5 * IQR, where IQR, the 'Inter Quartile Range', is the difference between the 1st and the 3rd quartile (= 25% and 75% percentile).
```{r 1.3}
library(dplyr)

is_outlier <- function(x) {
  return(x < quantile(x, 0.25) - 1.5 * IQR(x) | x > quantile(x, 0.75) + 1.5 * IQR(x))
}

shva %>%
  group_by(vowel) %>%
  mutate(outlier = ifelse(is_outlier(f1), f1, as.numeric(NA))) %>%
  ggplot(., aes(x = factor(vowel), y = f1)) +
    geom_boxplot() +
    geom_text(aes(label = outlier), na.rm = TRUE, hjust = -0.3)

print("679, 686, 826")
```

### 1.4 Calculate Pearson's correlation of `f1` and `f2` (all data)
```{r 1.4}
cor(shva[,4:5], method="pearson")
```

### 1.5 Calculate Pearson's correlation of `f1` and `f2` for each vowel
```{r 1.5}
a_shva <- shva %>% 
filter(shva$vowel=="a") 
cor(a_shva[,4:5],method='pearson')

A_shva <- shva %>% 
filter(shva$vowel=="A") 
cor(A_shva[,4:5],method='pearson')

y_shva <- shva %>% 
filter(shva$vowel=="y") 
cor(y_shva[,4:5],method='pearson')
```

### 1.6 Use the linear regression model to predict f2 by f1.
### 1.6.1 Provide the result regression formula
```{r}
fit1 = lm(f1~f2, data=shva)
summary(fit1)
```
### 1.6.2 Provide the adjusted R2
```{r}
print("Adjusted R-squared:  0.3319")
```
### 1.6.3 Add the regression line in scatterplot 1.1
```{r}
ggplot(shva, aes(x=f2, y=f1, color=vowel)) + geom_point() + scale_y_continuous(trans = "reverse") + scale_x_continuous(trans = "reverse") + geom_smooth(method='lm', se=FALSE, color='gray', size=0.5) + theme(legend.position="none") + ggtitle("f2 and f1 of the reduced and stressed vowels") + labs(caption="Data from Duryagin 2018")
```

### 1.7 Use the mixed-efects model to predict f2 by f1 using vowel intercept as a random effect

### 1.7.1 Provide the fixed effects formula
```{r}
rlm = lmer(f2~f1 + (1|vowel), data=shva)
summary(rlm)
fixef(rlm)
```

### 1.7.2 Provide the variance for intercept argument for vowel random effects
```{r}
print("Variance equals to 6374")
```

### 1.7.3 Add the regression line in scatterplot 1.1
```{r}
rlm2 = lmer(f1~f2 + (1|vowel), data=shva)
shva$random <- predict(rlm2)
ggplot(data = shva, aes(y = f1, x = f2, color=vowel)) +
  geom_point() + ggtitle("f1 and f2 of the reduced and stressed vowels") + 
  scale_y_reverse() + scale_x_reverse() +   theme(legend.position="none") + labs(caption = "Data from Duryagin 2018") +
geom_line(aes(f2, random, color = vowel))
```


## 2. English Lexicon Project data
880 nouns, adjectives and verbs from the English Lexicon Project data (Balota et al. 2007).

* `Format` -- A data frame with 880 observations on the following 5 variables.
* `Word` -- a factor with lexical stimuli.
* `Length` -- a numeric vector with word lengths.
* `SUBTLWF` -- a numeric vector with frequencies in film subtitles.
* `POS` -- a factor with levels JJ (adjective) NN (noun) VB (verb)
* `Mean_RT` -- a numeric vector with mean reaction times in a lexical decision task
Source (http://elexicon.wustl.edu/WordStart.asp)

Data from Natalya Levshina's `RLing` package available (here)[https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/data/ELP.csv]

### 2.0 Read the data from file to the variable `elp`.
```{r 2.0}
elp = read.csv("https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/data/ELP.csv")
```

### 2.1 Which two variables have the highest Pearson's correlaton value?
```{r 2.1}
cor1 = cor(elp[,2:3], method="pearson")
cor1

cor2 = cor(elp[,c(2,5)], method="pearson")
cor2

cor3 = cor(elp[,c(3,5)], method="pearson")
cor3

print("Length and Mean_RT have the highest Pearson's correlation value equal to 0.5276826")
```

### 2.2 Group your data by parts of speech and make a scatterplot of SUBTLWF and Mean_RT.
```{r 2.2}
by_POS = elp %>% group_by(POS)
by_POS

noun = elp %>% filter(elp$POS=="NN")
verb = elp %>% filter(elp$POS=="VB")
adj = elp %>% filter(elp$POS=="JJ")

elp %>%
  ggplot(aes(SUBTLWF, Mean_RT, color=Length)) + geom_point() + facet_wrap( ~POS, ncol=3) + scale_color_continuous(low = "lightblue", high = "red") + xlim(0,170) + scale_x_log10() + labs(caption="Data from (Balota et al. 2007)") + theme_bw()

```

We've used `scale_color_continuous(low = "lightblue", high = "red")` as a parameter of `ggplot()`.

### 2.3 Use the linear regression model to predict `Mean_RT` by `log(SUBTLWF)` and `POS`.
#### 2.3.1 Provide the result regression formula
```{r 2.3.1}
fit = lm(Mean_RT ~ log(SUBTLWF)+POS, data=elp)
summary(fit)
```

#### 2.3.2 Provide the adjusted R$^2$
```{r 2.3.2}
print("Adjusted R-squared: 0.3385")
```

#### 2.3.3 Add the regression line in the scatterplot.
```{r 2.3.3}
elp %>%
  ggplot(aes(log(SUBTLWF), Mean_RT, color=Length)) + geom_point() + scale_color_continuous(low = "lightblue", high = "red") + geom_smooth(method='lm', se=FALSE, color='black', size=0.5) + theme_bw() + labs(caption="Data from (Balota et al. 2007)")
```

### 2.4 Use the mixed-efects model to predict `Mean_RT` by `log(SUBTLWF)` using POS intercept as a random effect

#### 2.4.1 Provide the fixed effects formula
```{r 2.4.1}
random = lmer(Mean_RT~log(SUBTLWF)+(1|POS), data = elp)
summary(random)
```

#### 2.4.2 Provide the variance for intercept argument for `POS` random effects
```{r 2.4.2}
print("The variance for intercept argument for `POS` random effects = 414.4")
```

#### 2.4.3 Add the regression line to the scatterplot
```{r 2.4.3}
elp$random = predict(random)

elp %>%
  ggplot(aes(log(SUBTLWF), Mean_RT, color=POS)) + geom_point() + geom_line(aes(y=random), color='black', size=0.5) + facet_wrap( ~POS, ncol=3) + xlim(-4,8) + scale_x_continuous(breaks=c(-4,0,4,8)) + labs(caption="Data from (Balota et al. 2007)")
```

## 3. Dutch causative constructions

This is a data set with examples of two Dutch periphrastic causatives extracted from newspaper corpora.

The data frame includes 100 observations on the following 7 variables:

* Cx -- a factor with levels doen_V and laten_V
* CrSem -- a factor that contains the semantic class of the Causer with levels Anim (animate) and Inanim (inanimate).
* CeSem -- a factor that describes the semantic class of the Causee with levels Anim (animate) and Inanim (inanimate).
* CdEv -- a factor that describes the semantic domain of the caused event expressed by the Effected Predicate. The levels are Ment (mental), Phys (physical) and Soc (social).
* Neg -- a factor with levels No (absence of negation) and Yes (presence of negation).
* Coref -- a factor with levels No (no coreferentiality) and Yes (coreferentiality).
* Poss -- a factor with levels No (no overt expression of possession) Yes (overt expression of possession)

Data from Natalya Levshina's `RLing` package available (here)[https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/data/dutch_causatives.csv]

### 3.0 Read the data from file to the variable `d_caus`.
```{r 3.0}
d_caus = read.csv("https://raw.githubusercontent.com/agricolamz/2018-MAG_R_course/master/data/dutch_causatives.csv")
d_caus
```

### 3.1 We are going to test whether the association between `Aux` and other categorical variables (`Aux` ~ `CrSem`, `Aux` ~ `CeSem`, etc) is statistically significant. The assiciation with which variable should be analysed using Fisher's Exact Test and not using Pearson's Chi-squared Test? Is this association statistically significant?
```{r 3.1}
#Aux~CrSem
table1=table(d_caus$Aux, d_caus$CrSem)
chisq.test(table1)
chisq.test(table1)$expected

#Aux~CeSem
table2=table(d_caus$Aux, d_caus$CeSem)
chisq.test(table2)
chisq.test(table2)$expected

#Aux~CdEvSem
table3=table(d_caus$Aux, d_caus$CdEvSem)
chisq.test(table3)
chisq.test(table3)$expected

#Aux~CeSynt
table4=table(d_caus$Aux, d_caus$CeSynt)
chisq.test(table4)$expected
fisher.test(table4)

#Aux~EPTrans
table5=table(d_caus$Aux, d_caus$EPTrans)
chisq.test(table5)
chisq.test(table5)$expected

#Aux~Country
table6=table(d_caus$Aux, d_caus$Country)
chisq.test(table6)
chisq.test(table6)$expected

#Aux~Domain
table7=table(d_caus$Aux, d_caus$Domain)
chisq.test(table7)
chisq.test(table7)$expected

print("Association with CeSynt should be analysed using Fisher's Exact Test because one of the expected frequencies is less than 5. This association is statistically significant since p-value < 2.2e-16.")
```

### 3.2. Test the hypothesis that `Aux` and `EPTrans` are not independent with the help of Pearson's Chi-squared Test. 
```{r 3.2}
chisq = chisq.test(table5)
chisq

print("As p-value 0.0001553 is smaller than the .001 significance level, I reject the null hypothesis")
```

### 3.3 Provide expected values for Pearson's Chi-squared Test of `Aux` and `EPTrans` variables.
```{r 3.3}
chisq.test(table5)$expected
```

### 3.4. Calculate the odds ratio.
```{r 3.4}
fit = glm(Aux~EPTrans, family = binomial, data = d_caus)
summary(fit)

levels(d_caus$Aux)
levels(d_caus$EPTrans)

exp(summary(fit)$coefficients["EPTransTr",1])
```

### 3.5 Calculate effect size for this test using Cramer's V (phi).
```{r 3.5}
newtable = xtabs(~Aux+EPTrans, data=d_caus)
newtable #to have a frequency table in front of my eyes

assocstats(table5)
```

### 3.6. Report the results of independence test using the following template:
```
We have found a significant association between variables Aux and EPTrans (p < 0.001).  The odds of the effected predicate being transitive were 2.6 times higher in 'laten' than in 'doen'. Effect size is small (Cramer's V = 0.174).
```

### 3.7 Visualize the distribution using mosaic plot.
Use `mosaic()` function from `vcd` library.
```{r 3.7}
mosaic(~ Aux + EPTrans, data=d_caus, shade=TRUE, legend=TRUE)
```

Below is an example of how to use mosaic() with three variables.
```{r 3.7.1}
# mosaic(~ Aux + CrSem + Country, data=d_caus, shade=TRUE, legend=TRUE)
```

### 3.8 Why is it not recommended to run multiple Chisq tests of independence on different variables within your dataset whithout adjusting for the multiplicity? (i.e. just testing all the pairs of variables one by one)  
```
Running multiple Chi-square tests on different variables in the dataset without adjusting for the multiplicity may result in the type 1 error. Null hypotheses may be falsely rejected because the groups are more likely to differ significantly in one or several attributes and we combine confidence intervals of multiple comparisons therefore raising the likelihood of type 1 error.
```

### 3.9 Provide a short text (300 words) describing the hypothesis on this study and the results of your analysis.
```
The dataset contains information about the distribution of Dutch causatives and certain parameters that may influence it, in particular animacy of a causer or causee, the semantic domain of the caused event, transitivity or intransitivity of the effected predicate, Netherlandic or Belgian variety of Dutch and the domain. The null hypothesis is that the distribution of Dutch causatives doen and laten ('Aux') does not depend on the transitivity of the effected predicate ('EPTrans'). With the help of a Chi-square test I rejected the null hypothesis, p-value being smaller than the .001 significance level. However, the effect is rather small, with Cramer's V at 0.174. What is more, the probability of laten is greater if the effected predicate in intransitive and doen is used more often if the predicate is intransitive.
Regarding the association between the distribution of causatives and other parameters, one of the associations has to be analysed using Fisher's exact test - the association between causatives and Causee syntactic position. This association has been discivered to be statistically significant, with p-value less than 2.2e-16. Other than that, parameters may be explored using Chi-square test. However, running multiple Chi-square tests without adjusting for the multiplicity is not advised since it heightens the likelihood of a type 1 error - that is, rejecting a null hypothesis when it is in fact true.
```
